Regression models for Cylindrical data in Psychology

### 2019-03-03 Manuscript/Manuscript.tex
Manuscript Update

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

@@

In social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., linear
component). Many established models in psychology are often referred to
as circular or circumplex models, but those models are cylindrical.
Examples of such cylindrical models are the interpersonal
circle/circumplex (Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans,
\DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006),
the circumplex of affect (Russell, 1980), the circumplex of human
emotion (Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline \indent Also, many of the more recent types of data that are
studied in psychology are cylindrical. For example, research on human
navigation uses data where distance (i.e., linear component) and
direction (i.e., the linear component) are of interest (Chrastil \&
Warren, 2017) or in eye-tracking, the saccade data also consist of both
the direction (i.e., the circular variable) and the duration (i.e., the
linear variable)(e.g., Rayner (2009)). Apart from the social sciences
data with a circular and linear outcome more commonly occur in
meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013),
ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \&
Pereira, 2014) or marine research (Lagona, Picone, Maruotti, \& Cosoli,
2015) \newline \indent Up until now researchers studying cylindrical
data had to rely on linear statistical methods to analyze their research
results. However, lately more and more of these researchers acknowledge
that linear methods are not sufficient and call for new methods
(Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth,
2009) that take into account both the circular and the linear component
of these data. \newline \indent The aim in the present paper is twofold.
Firstly, we we intend to fill the above mentioned gap in the literature
by showing the use of cylindrical models can benefit the analysis of
circumplex data and cylindrical data in psychology in general. More
specifically we will show these benefits for interpersonal teacher data
from the field of educational psychology. Apart from modelling the
dependence between the linear and circular component of a cylindrical
variable we would also like to predict the two components from a set of
covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a
regression of both the linear and circular component of a cylindrical
variable onto a set of covariates. From now on we will therefore refer
to the components of the cylindrical variable as outcome components.
These adapted circular models are then used to analyse the teacher data.

@@

Data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear
variable in the sense that it is measured on a different scale. Figure
\ref{circline} shows the difference between a circular scale (right) and
a linear scale (left). The most important difference is that on a
circular scale the datapoints 0\(^\circ\) and 360\(^\circ\) are
connected and in fact represent the same number while on a linear scale
the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on
different places on the scale. Both circular data and cylindrical data
require special analysis methods due to this periodicity in the scale of
a circular variable (see e.g.~Fisher (1995) for an introduction to
circular data and Mardia and Jupp (2000), Jammalamadaka and Sengupta
(2001) and Ley and Verdebout (2017) for a more elaborate
overview).\newline \indent In the literature, several methods have been
put forward to model the relation between the linear and circular
component of a cylindrical variable. Some of these are based on
regressing the linear component onto the circular component using the
following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe and Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
respective shape and concentration parameters. However, none of the
methods that have been proposed thus far include additional covariates
onto which both the circular and linear component are regressed.

@@

The motivating example for this article comes from the field of
educational psychology and was collected for the studies on classroom
climate of \VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016)
and Pennings (2017a). An indicator of the quality of the classroom
climate is the students' perception of their teachers' interpersonal
behavior. These interpersonal perceptions, both in educational
psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz and Strack (2011)
for an overview of many such instruments).\newline \indent The
circumplex data used in this paper are measured using the Questionnaire
on Teacher Interaction (QTI) (Wubbels et al., 2006) which is one such
circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and
contains items that load on two interpersonal dimensions: Agency and
Communion. Agency refers to the degree of power or control a teacher
exerts in interaction with his/her students. Communion refers to the
degree of friendliness or affiliation a teacher conveys in interaction
with his/her students. The loadings on the two dimensions of the QTI can
be placed in a two-dimensional space formed by Agency (vertical) and
Communion (horizontal), see Figure \ref{QTI}. This space is called the
interpersonal circle/circumplex (IPC) and different parts of this space
are characterized by different teacher behavior, e.g.~\enquote{helpful}
or \enquote{uncertain}. The IPC is ``a continuous order with no
beginning or end'' (Gurtman, 2009, p. 2). We call such ordering a
circumplex ordering and the IPC is therefore often called the
interpersonal circumplex. The ordering also implies that scores on the
IPC could be viewed as a circular variable. This circular variable
represents the type of interpersonal behavior that a teacher shows
towards his/her students.\newline \indent Cremers et al. (2018a) explain
the circular nature of the IPC data and analyze them as such using a
circular regression model. The two dimension scores, Agency and
Communion, can be converted to a circular score using the two-argument
arctangent function in \eqref{PredVal}\footnote{The selection of

@@

bivariate space.}, where \(A\) represents a score on the Agency
dimension and \(C\) represents a score on the Communion dimension. Note
that when placing a unit circle on Figure \ref{QTI} we see that the
Agency dimension is related to the sine of the circular score and the
Communion dimension is related to the cosine of the circular score.

@@

\end{equation} The resulting circular variable \(\theta\) can then be
modelled and takes values in the interval \([0, 2\pi)\) or
\([0^\circ, 360^\circ)\). Note that the round brackets mean that
\(2\pi\) and \(360^\circ\) are not included in the interval since these
represent the same value as 0 as a result of periodicity. \newline
\indent A circular analysis of circumplex data has several benefits: it
is more in line with its theoretical definition and it allows us to
analyse the blend of the two dimensions Agency and Communion instead of
both dimensions separately. This provides us with new insights compared
to a separate analysis of the two dimensions that is standard in the
literature (see e.g. Pennings et al. (2018), Wright et al. (2009) or

@@

two-dimensional data are converted to the circle we lose some
information, namely the length of the two-dimensional vector
\((A, C)^t\), \emph{i.e.}, its Euclidean norm
\(\mid\mid (A, C)^t \mid\mid\). This length represents the strength of
the interpersonal behavior a teacher shows towards his/her students. In
a cylindrical model this strength (the linear outcome) can be modeled
together with the type of interpersonal behavior of a teacher (the
circular outcome). This leads to an improved analysis of interpersonal
circumplex data, over either analyzing the two dimensions separately or
using a circular model, because we take all information, circular and
linear, into account. In the next section we introduce several
cylindrical models that can be used to analyze the teacher data. First
however we will provide descriptives for the teacher data.

@@

The teacher data was collected between 2010 and 2015 and contains
several repeated measures on the IPC of 161 teachers. Measurements were
obtained using the QTI and taken in different years and classes. For
this paper we only consider one measurement, the first occasion (2010)
and largest class if data for multiple classes were available. This
results in a sample of 151 teachers. In addition to the type of
interpersonal behavior (IPC), the circular outcome, and the strength of
interpersonal behavior (IPC strength), the linear outcome, a teachers'
self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. In previous research, in psychology and
education it has been shown that higher self-efficacy is related to the
quality of interpersonal interactions (Locke \& Sadler, 2007;
\VANDER{Want}{Van der}{van der} Want et al., 2018). After listwise
deletion of missings (\(3\) in total, only for the self-efficacy) we
have a sample of 148 teachers. Table \ref{Tableteacherdescriptives}
shows descriptives for the dataset. For the circular variable IPC we
show sample estimates for the circular mean \(\bar{\theta}\) and
concentration \(\hat{\rho}\). The circular concentration lies between 0,
meaning the data is not concentrated at all \emph{i.e.} spread over the
entire circle, and 1, meaning all data is concentrated at a single point
on the circle The population values of these parameters are usually, and
also in this paper, referred to as \(\mu\) (circular location) and
\(\kappa\) (circular concentration). For the linear variables (strength
IPC and SE) we show sample estimates of the linear mean and standard
deviation (sd). Figure \ref{dataplot} is a scatterplot showing the
relation between the linear and circular outcome of the teacher data for
teachers with low SE (below 1 sd below the mean), average SE (between 1
sd below and 1 sd above the mean) and high SE (above 1 sd above the
mean).

@@

One of the goals of this paper is to show the benefits of cylindrical
methods for the analysis of circumplex data and cylindrical data in
psychology in general. To do so we decide to focus on four cylindrical
models. The models were selected for their relatively low complexity and
the ease with which a regression structure could be incorporated. But
also because they show different ways of modelling the the linear and
circular outcome and thereby illustrate a wider range of cylindrical
models available in the literature. The cylindrical models contain a set
of \(q\) predictors \(\boldsymbol{x} = x_1, \dots, x_q\) and \(p\)
predictors \(\boldsymbol{z} = z_1, \dots, z_q\) for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two
models are based on a construction by Mastrantonio et al. (2015), while
the other models are extensions of the models from Abe and Ley (2017)
and Mastrantonio (2018). The four cylindrical models are introduced
separately in the subsections below. However, to provide a more succinct
overview and comparison of the four models, Table \ref{TableModels}
gives an overview of the similarities and differences between the
models.

\begin{sidewaystable}[h]
\centering
\caption{Comparison of the four cylindrical regression models} 
\begin{tabular}{lllll}
  \noalign{\smallskip}\hline\noalign{\smallskip}
\multicolumn{1}{l}{Aspect} & CL-PN & CL-GPN  & Abe-Ley  & GPN-SSN \\ \hline\noalign{\smallskip}
$\Theta$ & &&&\\
$\:\:$Distribution& PN & GPN & Sine-skewed vM & GPN\\
$\:\:$Domain & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$\\
$\:\:$Shape & symmetric, & assymetric, & assymetric, & assymetric, \\
            & unimodal  & multimodal & unimodal   & multimodal \\\hline\noalign{\smallskip}
$Y$& &&&\\
$\:\:$Distribution & Normal & Normal & Weibull & skewed-Normal\\
$\:\:$Domain & $(-\infty, + \infty)$ & $(-\infty, + \infty)$ & $(0, + \infty)$ & $(0, + \infty)$\\
$\:\:$Shape & symmetric, & symmetric, & assymetric, & assymetric, \\
            & unimodal  & unimodal  & unimodal   & unimodal\\\hline\noalign{\smallskip}
$\Theta$-$Y$ dependence &                                   &                                   & & \\
                        & $y$ regressed on                  & $y$ regressed on                  & $\alpha$ and $\kappa$ & multivariate \\
                        & $\sin(\theta)$ and $\cos(\theta)$ & $\sin(\theta)$ and $\cos(\theta)$ & & distribution\\\hline
\multicolumn{5}{l}{Note: PN and GPN refer to the projected normal and general projected normal distribution.}\\
\multicolumn{5}{l}{vM refers to the von-Mises distribution}\\

@@

Following Mastrantonio et al. (2015) we consider two models where the
relation between \(\Theta \in [0, 2\pi)\) and
\(Y\in (-\infty, + \infty)\) and \(q\) covariates is specified as

@@

\end{equation} \noindent where the random variable \(R\geq0\) will be
introduced below, the error term \(\epsilon \sim N(0, \sigma^2)\) with
variance \(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(x_1, \dots, x_q\) are
the \(q\) covariates for the prediction of the linear outcome. We thus
assume a normal distribution for the linear outcome.\newline \indent For
the circular outcome we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise
from a projection of a distribution defined in bivariate space onto the
circle. Figure \ref{projection} represents this projection. In the left
plot of Figure \ref{projection} we see datapoints from the bivariate
variable \(\boldsymbol{S}\) that in the middle plot are projected to
form the circular outcome \(\Theta\) in the right plot. Mathematically
the relation between \(\boldsymbol{S}\) and \(\Theta\) is defined as
follows \begin{equation}\label{projection}

@@

\end{equation} \noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\),
the Euclidean norm of \(\boldsymbol{S}\); the lines connecting the
bivariate datapoints to the origin in the middle plot. We call
\(\boldsymbol{S}\) the augmented representation of the circular outcome,
it is a variable that in contrast to \(\Theta\) is not observed and thus
considered latent or auxiliary. This then means that we do not model
\(\Theta\) directly but indirectly through \(\boldsymbol{S}\). \newline
\indent For both the PN and GPN distribution the circular location
parameter \(\mu\in [0, 2\pi)\) is modeled as
\(\hat{\mu}_i = \mbox{atan2}(\hat{\mu}_i^{II}, \hat{\mu}_i^{I}) = \mbox{atan2}(\boldsymbol{\beta}^{II}\boldsymbol{z}_i, \boldsymbol{\beta}^{I}\boldsymbol{z}_i)\)\footnote{Note that for the CL-GPN

@@

coefficients for the prediction of \(\boldsymbol{S}\) and
\(\boldsymbol{z}_i\) is a vector with predictor values for each
individual \(i \in 1, \dots, n\) where \(n\) is the sample size. Note
that as a result of the augmented representation of the circular outcome
we have two sets of regression coefficients and intercepts, one for each
bivariate component of \(\boldsymbol{S}\). This leads to problems when
we want to interpret the effect of a covariate on the circle. A circular
regression line is shown in Figure \ref{circregline}, with covariate
values on the x-axis and the predicted circular outcome on the y-axis.
As can be seen it is of a non-linear character meaning that the effect
of a covariate is different at different values of the covariate. A
circular regression line is usually described by the slope at the
inflection point, the point at which the slope of the regression line
starts flattening off (indicated with a square in Figure
\ref{circregline}). By default, the parameters from the PN and GPN model
do not directly describe this inflection point. For the PN distribution
however, Cremers et al. (2018b) solved this interpretation problem and
introduce new circular regression coefficients. They introduce a new
parameter \(b_c\) that describes the slope at the inflection point of
the regression line. For the GPN distrbution the interpretation problem
however remains.\newline \indent The main difference between the PN and
GPN distribution lies in the definition of their covariance matrix. For
the PN distribution this is an identity matrix, causing the distribution
to be unimodal and symmetric, whereas for the GPN distribution
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \xi^2 & \xi\\ \xi & 1 \end{bmatrix}\),
allowing for multimodality and assymetry/skewness.\newline \indent For
the teacher dataset the predicted linear outcome, strengths of

@@

\gamma_{sin}\sin(\theta_i)r_i + \gamma_1\text{SE}_i.\] The predicted
circular outcome, type of interpersonal behavior, equals:
\[\hat{\theta}_i = \mu_i =

@@

\beta_1^{I}\text{SE}_i).\] \noindent where \(\text{SE}_i\) is the
self-efficacy score of one individual. The CL-PN and CL-GPN models thus
allow us to assess the average type and strength of interpersonal
behavior through the parameters \(\beta_{0}^{I}\), \(\beta_{0}^{II}\)
and \(\gamma_0\) as well as the effect of self-efficacy on type and
strength of teacher behavior through the parameters, \(\beta_{1}^{I}\),
\(\beta_{1}^{II}\) and \(\gamma_1\). In addition, because the type and
strength of interpersonal behavior are modelled together via the
regression in \eqref{circlinlink} we can assess the effect of the type
of interpersonal behavior on the strength through the parameters
\(\gamma_{sin}\) and \(\gamma_{cos}\). In the teacher data these are the
regression coefficients for the effect of the sine and cosine of the
type of behavior which are related to the scores on the Agency an
Communion dimensions respectively.\newline \indent Both the CL-PN and
CL-GPN models are estimated using Markov Chain Monte Carlo (MCMC)
methods based on Nuñez-Antonio, Gutiérrez-Peña, and Escarela (2011),
Wang and Gelfand (2013) and Hernandez-Stumpfhauser, Breidt, and Woerd
(2016) for the regression of the circular outcome. A detailed
description of the Bayesian estimation and MCMC samplers can be found in
the Supplementary Material.

@@

The parameters \(\gamma_1\) in the CL-PN, CL-GPN, Abe-Ley model and
\(\beta_{1_y}\) in the GPN-SSN model inform us about the effect of
self-efficacy on the strength of interpersonal behavior. For the CL-PN,

>>>>

\begin{document}
\maketitle

@@

In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. Gurtman (2011) refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., the linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
(Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans, \DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006), the circumplex of
affect (Russell, 1980), the circumplex of human emotion
(Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., linear component) and direction (i.e., the circular component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable) (e.g., Rayner (2009)). Apart from the social
sciences, data with a circular and linear outcome more commonly occur in
meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \& Pereira, 2014) or
marine research (Lagona, Picone, Maruotti, \& Cosoli, 2015)
\newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill
the above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. These adapted cylindrical models are
then used to analyse the teacher data.

@@

Data that consist of a linear variable and a circular variable is called
cylindrical data. A circular variable is different from the linear variable in
the sense that it is measured on a different scale. Figure \ref{circline} shows
the difference between a circular scale (right) and a linear scale (left). The
most important difference is that on a circular scale the datapoints 0\(^\circ\)
and 360\(^\circ\) are connected and in fact represent the same number while on a
linear scale the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on different
places on the scale. Both circular data and cylindrical data require special
analysis methods due to this periodicity in the scale of a circular variable
(see e.g.~Fisher (1995) for an introduction to circular data and
Mardia and Jupp (2000), Jammalamadaka and Sengupta (2001) and Ley and Verdebout (2017) for a more
elaborate overview).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component (Johnson \& Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \& Jona-Lasinio, 2015). Others model the relation in a
different way, e.g.~by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
(Mastrantonio, 2018) or by proposing a joint cylindrical distribution. For
example, Abe and Ley (2017) introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.

@@

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
\VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016) and Pennings (2017a). An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz and Strack (2011) for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) (Wubbels et al., 2006) which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. Agency refers
to the degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. This space is
called the interpersonal circle/circumplex (IPC) and different parts of this
space are characterized by different teacher behavior, e.g.~\enquote{helpful} or
\enquote{uncertain}. The IPC is ``a continuous order with no beginning or end''
(Gurtman, 2009, p. 2). We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable. This
circular variable represents the type of interpersonal behavior that a teacher
shows towards his/her students.\newline
\indent Cremers et al. (2018a) explain the circular nature of the IPC data and
analyze them as such using a circular regression model. The two dimension
scores, Agency and Communion, can be converted to a circular score using the
two-argument arctangent function in \eqref{PredVal}, where \(A\) represents a score on the Agency dimension and
\(C\) represents a score on the Communion dimension \footnote{The selection of

@@

bivariate space.}. Note that when placing a
unit circle on Figure \ref{QTI} we see that the Agency dimension is related to
the sine of the circular score and the Communion dimension is related to the
cosine of the circular score.

@@

\end{equation}
The resulting circular variable \(\theta\) can then be modelled and takes values
in the interval \([0, 2\pi)\) or \([0^\circ, 360^\circ)\). Note that the round
brackets mean that \(2\pi\) and \(360^\circ\) are not included in the interval since
these represent the same value as 0 as a result of periodicity. \newline
\indent A circular analysis of circumplex data has several benefits: it is more
in line with its theoretical definition and it allows us to analyse the blend of
the two dimensions Agency and Communion instead of both dimensions separately.
This provides us with new insights compared to a separate analysis of the two
dimensions that is standard in the literature (see e.g.
Pennings et al. (2018), Wright et al. (2009) or

@@

two-dimensional data are converted to the circle we lose some information,
namely the length of the two-dimensional vector \((A, C)^t\), \emph{i.e.}, its
Euclidean norm \(\mid\mid (A, C)^t \mid\mid\). This length represents the
strength of the interpersonal behavior a teacher shows towards his/her students.
In a cylindrical model this strength (the linear outcome) can be modeled
together with the type of interpersonal behavior of a teacher (the circular
outcome). This leads to an improved analysis of interpersonal circumplex data,
over either analyzing the two dimensions separately or using a circular model,
because we take all information, circular and linear, into account. In the next
section we introduce several cylindrical models that can be used to analyze the
teacher data. First however we will provide descriptives for the teacher data.

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. This results in a sample of 151 teachers.
In addition to the type of interpersonal behavior (IPC), the circular outcome,
and the strength of interpersonal behavior (IPC strength), the linear outcome, a
teachers' self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. In previous research, in psychology and education it
has been shown that higher self-efficacy is related to the quality of
interpersonal interactions (Locke \& Sadler, 2007; \VANDER{Want}{Van der}{van der} Want et al., 2018).
After listwise deletion of missings (\(3\) in total, only for the self-efficacy)
we have a sample of 148 teachers. Table \ref{Tableteacherdescriptives} shows
descriptives for the dataset. For the circular variable IPC we show sample
estimates for the circular mean \(\bar{\theta}\) and concentration \(\hat{\rho}\).
The circular concentration lies between 0, meaning the data is not concentrated
at all \emph{i.e.} spread over the entire circle, and 1, meaning all data is
concentrated at a single point on the circle. The population values of these
parameters are usually, and also in this paper, referred to as \(\mu\) (circular
location) and \(\kappa\) (circular concentration). For the linear variables
(strength IPC and SE) we show sample estimates of the linear mean and standard
deviation (sd). Figure \ref{dataplot} is a scatterplot showing the relation between
the linear and circular outcome of the teacher data for teachers with low SE
(below 1 sd below the mean), average SE (between 1 sd below and 1 sd above the
mean) and high SE (above 1 sd above the mean).

@@

One of the goals of this paper is to show the benefits of cylindrical methods
for the analysis of circumplex data and cylindrical data in psychology in
general. To do so we decide to focus on four cylindrical models. The models were
selected for their relatively low complexity and the ease with which a
regression structure could be incorporated. But also because they show different
ways of modelling the linear and circular outcome and thereby illustrate a
wider range of cylindrical models available in the literature. The
cylindrical models contain a set of \(q\) predictors \(\boldsymbol{x} = x_1, \dots, x_q\) and \(p\) predictors \(\boldsymbol{z} = z_1, \dots, z_p\) for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two models are
based on a construction by Mastrantonio et al. (2015), while the other models are
extensions of the models from Abe and Ley (2017) and Mastrantonio (2018). The
four cylindrical models are introduced separately in the subsections below.
However, to provide a more succinct overview and comparison of the four models, Table
\ref{TableModels} gives an overview of the similarities and differences between
the models.

\begin{sidewaystable}[h]
\centering
\caption{Comparison of the four cylindrical regression models} 
\begin{tabular}{lllll}
  \noalign{\smallskip}\hline\noalign{\smallskip}
\multicolumn{1}{l}{Aspect} & CL-PN & CL-GPN  & Abe-Ley  & GPN-SSN \\ \hline\noalign{\smallskip}
$\Theta$ & &&&\\
$\:\:$Distribution& PN & GPN & Sine-skewed vM & GPN\\
$\:\:$Domain & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$\\
$\:\:$Shape & symmetric, & asymmetric, & asymmetric, & asymmetric, \\
            & unimodal  & multimodal & unimodal   & multimodal \\\hline\noalign{\smallskip}
$Y$& &&&\\
$\:\:$Distribution & Normal & Normal & Weibull & skewed-Normal\\
$\:\:$Domain & $(-\infty, + \infty)$ & $(-\infty, + \infty)$ & $(0, + \infty)$ & $(-\infty, + \infty)$\\
$\:\:$Shape & symmetric, & symmetric, & asymmetric, & asymmetric, \\
            & unimodal  & unimodal  & unimodal   & unimodal\\\hline\noalign{\smallskip}
$\Theta$-$Y$ dependence &                                   &                                   & & \\
                        & $y$ regressed on                  & $y$ regressed on                  & $\alpha$ and $\kappa$ & multivariate \\
                        & $\sin(\theta)$ and $\cos(\theta)$ & $\sin(\theta)$ and $\cos(\theta)$ & & distribution\\\hline
\multicolumn{5}{l}{Note: PN and GPN refer to the projected normal and general projected normal distribution.}\\
\multicolumn{5}{l}{vM refers to the von-Mises distribution}\\

@@

Following Mastrantonio et al. (2015) we consider two models where
the relation between \(\Theta \in [0, 2\pi)\) and \(Y\in (-\infty, + \infty)\)
and \(q\) covariates is specified as

@@

\end{equation}
\noindent where the random variable \(R\geq0\) will be introduced below, the
error term \(\epsilon \sim N(0, \sigma^2)\) with variance \(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are the
intercept and regression coefficients and \(x_1, \dots, x_q\) are the \(q\)
covariates for the prediction of the linear outcome. We thus assume a normal
distribution for the linear outcome.\newline
\indent For the circular outcome we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise from a
projection of a distribution defined in bivariate space onto the circle. Figure
\ref{projection} represents this projection. In the left plot of Figure
\ref{projection} we see datapoints from the bivariate variable
\(\boldsymbol{S}\) that in the middle plot are projected to form the circular
outcome \(\Theta\) in the right plot. Mathematically the relation between
\(\boldsymbol{S}\) and \(\Theta\) is defined as follows
\begin{equation}\label{projection}

@@

\end{equation}
\noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\), the Euclidean norm of
\(\boldsymbol{S}\); the lines connecting the bivariate datapoints to the origin
in the middle plot. We call \(\boldsymbol{S}\) the augmented representation of the
circular outcome. It is a variable that in contrast to \(\Theta\) is not
observed and thus considered latent or auxiliary. This then means that we do not
model \(\Theta\) directly but indirectly through \(\boldsymbol{S}\). \newline
\indent For both the PN and GPN distributions the circular location parameter
\(\mu\in [0, 2\pi)\) is modeled as \(\hat{\mu}_i = \mbox{atan2}(\hat{\mu}_i^{II}, \hat{\mu}_i^{I}) = \mbox{atan2}(\boldsymbol{\beta}^{II}\boldsymbol{z}_i, \boldsymbol{\beta}^{I}\boldsymbol{z}_i)\)\footnote{Note that for the CL-GPN

@@

coefficients for the prediction of \(\boldsymbol{S}\) and \(\boldsymbol{z}_i\) is a
vector with predictor values for each individual \(i \in 1, \dots, n\) where \(n\)
is the sample size. Note that as a result of the augmented representation of the
circular outcome we have two sets of regression coefficients and intercepts, one
for each bivariate component of \(\boldsymbol{S}\). This leads to problems when we
want to interpret the effect of a covariate on the circle. A circular regression
line is shown in Figure \ref{circregline}, with covariate values on the x-axis
and the predicted circular outcome on the y-axis. As can be seen it is of a
non-linear character meaning that the effect of a covariate is different at
different values of the covariate. A circular regression line is usually
described by the slope at the inflection point, the point at which the slope of
the regression line starts flattening off (indicated with a square in Figure
\ref{circregline}). By default, the parameters from the PN and GPN models do not
directly describe this inflection point. For the PN distribution however,
Cremers et al. (2018b) solved this interpretation problem and introduce new
circular regression coefficients. They introduce a new parameter \(b_c\) that
describes the slope at the inflection point of the regression line. For the
GPN distrbution the interpretation problem however remains.\newline
\indent The main difference between the PN and GPN distribution lies in the
definition of their covariance matrix. For the PN distribution this is an
identity matrix, causing the distribution to be unimodal and symmetric, whereas
for the GPN distribution \(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \xi^2 & \xi\\ \xi & 1 \end{bmatrix}\) where \(\xi,\tau \in (-\infty, +\infty)\), allowing for multimodality and
asymetry/skewness.\newline
\indent For the teacher dataset the predicted linear outcome, strengths of

@@

\gamma_{sin}\sin(\theta_i)r_i + \gamma_1\text{SE}_i.\] The predicted circular
outcome, type of interpersonal behavior, equals: \[\hat{\theta}_i = \mu_i =

@@

\beta_1^{I}\text{SE}_i).\] \noindent where \(\text{SE}_i\) is the self-efficacy
score of one individual. The CL-PN and CL-GPN models thus allow us to assess the
average type and strength of interpersonal behavior through the parameters
\(\beta_{0}^{I}\), \(\beta_{0}^{II}\) and \(\gamma_0\) as well as the effect of
self-efficacy on type and strength of teacher behavior through the parameters,
\(\beta_{1}^{I}\), \(\beta_{1}^{II}\) and \(\gamma_1\). In addition, because the type
and strength of interpersonal behavior are modelled together via the regression
in \eqref{circlinlink} we can assess the effect of the type of interpersonal
behavior on the strength through the parameters \(\gamma_{sin}\) and
\(\gamma_{cos}\). In the teacher data these are the regression coefficients for
the effect of the sine and cosine of the type of behavior which are related to
the scores on the Agency and Communion dimensions respectively.\newline
\indent Both the CL-PN and CL-GPN models are estimated using Markov Chain Monte
Carlo (MCMC) methods based on Nuñez-Antonio, Gutiérrez-Peña, and Escarela (2011), Wang and Gelfand (2013) and
Hernandez-Stumpfhauser, Breidt, and Woerd (2016) for the regression of the circular outcome. A detailed
description of the Bayesian estimation and MCMC samplers can be found in the
Supplementary Material.

@@

The parameters \(\gamma_1\) in the CL-PN, CL-GPN, Abe-Ley models and \(\beta_{1_y}\)
in the GPN-SSN model inform us about the effect of self-efficacy on the strength
of interpersonal behavior. For the CL-PN, CL-GPN and GPN-SSN model the
